Extended Data

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
read/ cite delpierre_temperate_2016
#116 , #115

@camerondow35 , FYI, delpierre_temperate_2016 is a very relevant and interesting review!

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**

@@

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021] (but see Refs. [@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers). 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021]. <!--this sentence is repetitive with with previous.-->

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig. 3).
We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

@@

<!-- The following two paragraphs--and two final sentences in the paragraph above-- need work (assigned to KAT). It may make most sense to condense into a single paragraph making the point that while we don't know where the C goes, we can rule out a substantial, sustained boost to the long-term sink of woody productivity-->
One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Extra C could also be saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014]. 
Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig. pos. correlations for 5/142 RP and 3/66 DP species-site combinations, Fig. `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates. 
Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.

>>>>

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021; @delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**

@@

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021; @delpierre_temperate_2016] (but see Refs. [@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers). 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021]. <!--this sentence is repetitive with with previous.-->

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014; @finzi_carbon_2020]. 
**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig. 3).**
**We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`).** <!--repeated sentence-->
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

@@

<!-- The following two paragraphs--and two final sentences in the paragraph above-- need work (assigned to KAT). It may make most sense to condense into a single paragraph making the point that while we don't know where the C goes, we can rule out a substantial, sustained boost to the long-term sink of woody productivity-->
One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006; @roibu_climatic_2020; @kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006; @roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Extra C could also be saved within trees as non-structural carbohydrates (NSC) and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013], potentially including an earlier start to growth[@fu_variation_2014]. 
Extension of our tree-ring analysis revealed weak correlation between April $T_{max}$ and growth the following year (sig. pos. correlations for 5/142 RP and 3/66 DP species-site combinations, Fig. `r EDF_tree_rings_TMAX_extended`), although predominantly positive (non-significant) correlations in RP species suggests that this dynamic may weakly influence their annual growth.
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC production, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates. 
Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.
